Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-21 paper/summary.tex
first rough draft - MKL

\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of

This non-monotonic behaviour of the vortex lifetime was also reported
by \cite{fu14}, who performed isothermal disc-planet 
simluations with different values of the disc aspect ratio.   
% Cooling rates with
% $\tilde\beta>5.0$ showed lifetimes that were well below the max
% lifetime. 
By analyzing the radial region around the vortex within $[2r_{h},10r_{h}]$
% {\bf define this region}
we find the 
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. This value of $h$ was also
found to maximise vortex lifetimes in the simulations of
\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
our simulations $h$ is a dynamical variable.

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have carried out numerical simulations of non-isothermal disc-planet interaction. 
Our simulations were customised to examine the effect of a finite cooling time on the stability of gaps 
opened by giant planets to the so-called vortex or Rossby wave instability. To do so, we 
included an energy equation with a cooling term that restores the initial disc temperature on a characteristic timescale $t_c$. 
We studied the evolution of the gap stability as a function of $t_c$. This is a natural extension to previous
studies of on gap stability, which employ locally or strictly isothermal equations of state.

We considered two types of numerical experiments. 
We first used disc-planet interaction to self-consistently set up
gap profiles, which were then perturbed and evolved without further the influence of the planet potential. This procedure  
isolates the effect of cooling on gap stability through the set up of the initial gap profile. We find that as the cooling time $t_c$ is increased,
the gaps became more stable, with lower growth rates of non-axisymmetric modes and the dominant azimuthal wavenumber also decreases. 
This is consistent with the notion that increasing $t_c$ leads to higher temperatures or $h$, which opposes gap-opening by the planet. This means that
the gaps opened by the planet in a disc with longer $t_c$ is smoother and therefore  more stable to the RWI.

In the second set of calculations, we included the planet potential throughout the simulations 
and examined the long-term evolution of the gap-edge vortex that develops from the RWI. The vortices to reach a
quasi-steady state lasting $O(10^3)$ orbits, during which its amplitude grows until it begins to induce shocks,  
afterwhich the vortex amplitudes decay. 
We observe vortex lifetimes to increase with increasing cooling time until
a cooling timescale on the order of the libration time at the gap edge {\bf (this statement assumes longest lifetime is $\tilde{\beta}=1$, so may change
depending on new definitions of lifetimes)}, beyond which vortex lifetimes shorten again.

Our results indicate there exists an optimum cooling timescale to maximise the lifetime of gap-edge vortices. We suggest the non-monotonic dependence
can be attributed to the time required for the vortex to grow to sufficient amplitude to induce shocks in the surrounding fluid, thereby losing energy and also smooth out the gap edge.  
For short cooling timescales, the planet is able to open a deeper gap which favours the RWI, leading to stronger vortices. However, for long 
cooling timescales, the vortex grow faster during the quasi-steady state {\bf check if statement true}. 
We suggest the latter is due to the RWI being favoured with increasing disc temperature, according to previous stability
calculations. These competing factors imply for both very short and long cooling timescales, the vortex reaches its maximum amplitude 
, shocks, and begins to decay, sooner than intermediate cooling timescales. 
%decay timescale depends on cooling time, but there seems to be a monotonic trend: longer cooling times, decay timescale is shorter, so this 
%doesn't contribute to non-monotonic part

We remark that a non-monotonic dependence of the vortex lifetime was also reported
by \cite{fu14}, who performed isothermal disc-planet simluations with different values of the 
disc aspect ratio. In their simulations the optimum aspect ratio is $h=0.06$. In our simulations,
the aspect ratio is a dynamical variable, but by analysing the region where the vortex is located ($r-r_p\in[2,10]r_h$), 
% {\bf define this region}
we find for $\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$ on average throughout its lifetime. 
{\bf if new definition of vortex lifetime gives other $\tilde{\beta}$ values as the maximum lifetime, then need to update this value of $h$}. 
Our result is therefore qualitatively consistent with \citep{fu14}. 
%This value of $h$ was also
%found to maximise vortex lifetimes in the simulations of
%\cite{fu14}. Note that $h$ is a fixed parameter in \cite{fu14} while in
%our simulations $h$ is a dynamical variable.

\subsection{Caveats and outlooks}
We discuss below several outstanding issues that needs to be addressed in future works:

\emph{Convergence.} Although lower-resolution simulations performed in the early stages of this project gave similar results 
(most importantly, the non-monotonic dependence of vortex lifetimes on the cooling timescale), we did find the lower-resolution simulations 
typically yield longer vortex lifetimes than that reported in this paper. This could be due to weaker RWI with low resolution. 
It will be neccessary to perform even higher resolution  simulations in order to obtain quantitatively converged vortex lifetimes.

\emph{Self-gravity.} We have ignored disc self-gravity in this study. Based on linear calculations, \cite{lovelace13} concluded self-gravity
to be important for the RWI when $Q<O(1/h)$, or $Q\lesssim 20$ for $h\sim0.05$, as was typially considered in this work. This suggests that 
self-gravity may affect vortex lifetimes even when the Toomre $Q$ parameter is not small. In particular, given that we observe
vortices can reach significant over-densities (up to almost an order of magnitude), it will be important to include disc self-gravity 
in the future. %self-gravitational collapse of vortices

\emph{Three-dimensional (3D) effects.} A vortex in a 3D disc may be subject to secondary instabilities 
that destroy them \citep{lesur09,railton14}. This may an important factor in determining gap-edge vortex lifetimes
in realistic discs. For example, if these secondary instabilities sets in before the vortex grows to sufficient amplitude to shock, 
then the dependence of the vortex lifetime on the cooling timescale will be that through the 3D instability (as opposed to the effect on the RWI itself).  
This problem needs to be clarified with full 3D disc-planet simulations.

%3d is challenging because high resolution needed in for 3d instabilities
%However, the role of these instabilities on gap-edge vortices, where disc-planet  
%interaction can maintain the RWI, has not been clarified.
